## Paradoxical Tumors in Penile Cancer

### Summary
The aim of this study is to report the clinicopathologic and outcome features of superficial high-grade and deep low-grade penile squamous cell carcinomas. For this, we used a retrospectively-colleted series of patients with penile cancer, in which we identified 41 cases corresponding to 12 superficial high-grade tumors and 29 deep low-grade tumors. As outcomes we evaluated inguinal lymph node status, presence of tumor relapse, final nodal status, and cancer-specific death. Follow-up ranged from 0.8 to 386.7 months (mean, 152.5 months; median, 157.3 months). Clinicopathologic features were similar between superficial high-grade and deep low-grade tumors, except for a tendendy (Fisher’s exact P = 0.057) of the former to include tumors with a verruciform pattern of growth. A significantly higher proportion of inguinal lymph node metastasis was found in superficial high-grade tumors compared to deep low-grade tumors (80% vs. 20% respectively, Fisher’s exact P = 0.02). No significant differences were found regarding tumor relapse (Fisher’s exact P = 0.52), final nodal status (Mantel-Cox’s P = 0.42), or cancer-related death (Mantel-Cox’s P = 0.52). Our findings suggest that patients with superficial high-grade tumors may be treated differently from patients with deep low-grade tumors, at least to control short-term local disease. Prophylactic inguinal lymphadenectomuy might be indicated in cases of superficial tumors with high-grade histology while in deeply invasive low-grade penile carcinomas a more conservative approach may be considered.

### Description of the repository
This repository contains the full statistical analysis of the dataset that was used for the article _"Clinicopathologic and Outcome Features of Superficial High-Grade and Deep Low-Grade Squamous Cell Carcinomas of the Penis"_. The article is currently under consideration of publication. This repository also contains the following files:

* The final PDF version of the [article](https://github.com/alcideschaux/Penis-Paradoxical/blob/master/Article/Paradoxical_Tumors.pdf), as submitted for consideration of publication
* The [BibTeX](https://github.com/alcideschaux/Penis-Paradoxical/blob/master/Article/References.bib) file containing all the references cited in the article
* The [R script](https://github.com/alcideschaux/Penis-Paradoxical/blob/master/Article/Paradoxical_Tumors.R) that was used for analyzing the dataset and write the article
* The [R Markdown](https://github.com/alcideschaux/Penis-Paradoxical/blob/master/README.Rmd) file used for this report
* The [figures](https://github.com/alcideschaux/Penis-Paradoxical/tree/master/figure) included in this repository in PNG format

Data were analyzed using [R](http://www.r-project.org) version 3.1.2 “Pumpkin Helmet” (R Foundation for Statistical Computing, Vienna, Austria). Results were written using RMarkDown inside [RStudio](http://www.rstudio.com) version 0.98.1102, powered by the [knitr](http://cran.r-project.org/web/packages/knitr/index.html) package version 1.9 by [Yihui Xie](http://yihui.name/knitr).

### Building the dataset for analysis
First we loaded the full dataset including 333 patiens with invasive penile squamous cell carcinoma.

```{r DataSet}
Data <- read.csv("Article/PenisSCC_333.csv")
```

The full dataset is available at http://dx.doi.org/10.6084/m9.figshare.1290997, which also contains the dataset's codebook. We excluded patients who were lost at follow up in `Outcome`.

```{r CaseSelection}
Data <- subset(Data, Outcome != "Lost at follow-up")
```

We then recoded 3 variables, creating new variables cancer-related death (`DOD`), final nodal status (`Final_Nodal`), and anatomical location (`Anatomical`). Positive cases for `Final_Nodal`where those with lymph node metastasis in the groin dissection, local relapse during follow-up, or unfavorable outcome, including alive with disease and death by cancer. We also reordered `Subtype`.

```{r RecodeVariables}
# Creating a new variable for cancer-related death
Data$DOD <- ifelse(Data$Outcome == "Died of cancer", c("Yes"), c("No"))
Data$DOD <- factor(Data$DOD)
# Creating a new variable for the final nodal status
## Positive cases where those with:
  # Nodal metastasis in lymphadenectomy
  Data$Final_Nodal <- ifelse(Data$Mets == "Yes", c("Positive"), c("Negative"))
  # Local relapse during follow-up
  Data$Final_Nodal[Data$Local == "Yes"] <- "Positive"
  # Unfavorable outcome, including alive with disease and death by cancer
  Data$Final_Nodal[Data$Outcome == "Alive with disease"] <- "Positive"
  Data$Final_Nodal[Data$Outcome == "Died of cancer"] <- "Positive"
  Data$Final_Nodal <- factor(Data$Final_Nodal)
# Creating a new variable for anatomical location
Data$Anatomical[Data$Glans == "Yes" &
  Data$Sulcus == "No" &
  Data$Foreskin == "No"] <- "Glans alone"
Data$Anatomical[Data$Glans == "Yes" &
  Data$Sulcus == "Yes" &
  Data$Foreskin == "No"] <- "Glans + Coronal sulcus"
Data$Anatomical[Data$Glans == "Yes" &
  Data$Sulcus == "Yes" &
  Data$Foreskin == "Yes"] <- "Glans + Coronal sulcus + Foreskin"
Data$Anatomical <- factor(Data$Anatomical,
  levels = c("Glans alone", "Glans + Coronal sulcus",
    "Glans + Coronal sulcus + Foreskin"))
# Releveling variables
Data$Subtype <- factor(Data$Subtype,
  levels = c("Usual", "Verrucous", "Papillary", "Warty", "Mixed"))
```

We finally searched the dataset for cases of superficial high-grade and deep low-grade tumors, using the following criteria:

* __Superficial High-Grade Tumors:__ Grade 3 tumors invading lamina propria or superficial corpus spongiosum (tumor thickness equals to or less than 5 mm).
* __Deep Low-Grade Tumors:__ Grade 1 tumors invading deep corpus spongiosum (tumor thickness equals to or greater than 10 mm) or corpus cavernosum, including tunica albuginea.

We then excluded the missing cases and dropped unused levels.

```{r Paradoxical}
# Superficial High-Grade Tumors
Data$Paradoxical[Data$Grade == "Grade 3" &
  Data$Level == "Lamina propria"] <- "Superficial High-Grade"
Data$Paradoxical[Data$Grade == "Grade 3" &
  Data$Level == "Corpus spongiosum" &
  Data$Thickness <= 5] <- "Superficial High-Grade"
# Deep Low-Grade Tumors
Data$Paradoxical[Data$Grade == "Grade 1" &
  Data$Level == "Corpus cavernosum"] <- "Deep Low-Grade"
Data$Paradoxical[Data$Grade == "Grade 1" &
  Data$Level == "Corpus spongiosum" &
  Data$Thickness >= 10] <- "Deep Low-Grade"
# Converting to Factor
Data$Paradoxical <- factor(Data$Paradoxical,
  levels = c("Superficial High-Grade", "Deep Low-Grade"))
# Excluding missing cases
Data <- Data[complete.cases(Data$Paradoxical), ]
# Dropping unused levels
Data <- droplevels(Data)
```

Data analysis was carried out on this dataset, using the [simpleR](https://github.com/alcideschaux/simpleR) package.
```{r simpleR}
library(simpleR)
```

### Methodology
Data analysis is divided in 4 sections, as it follows:

__<a href="#Descriptive">Descriptive Statistics.</a>__ All the variables included in the dataset were analyzed using bar plots, histograms, box plots, and one-way tables. Factor variables were described using absolute and relative percentages. Numeric variables were described using mean, standard deviation, median, interquartile range, minimum and maximum value.

__<a href="#Inferential">Inferential Statistics:</a>__ Statistical tests (Fisher's exact test for categorical variables, Kruskal-Wallis test for numerical variables) were carried out to compare the clinicopathologic and outcome features between superficial high-grade and deep low-grade tumors. A 2-tailed P value was reported in all instances. Reported statistics included absolute and relative percentages for categorical variables; and mean, standard deviation, median, interquartile range, minimum and maximum value for numeric variables, by tumor type (superficial high-grade vs. deep low-grade).

__Survival Analysis.__ For all variables in the dataset survival curves were built for <a href="#Survival_FN">final nodal status</a> and <a href="#Survival_DOD">cancer-related death</a> using the Kaplan-Meier method. Survival curves were compared using the Mantel-Cox (log-rank) test. Numerical variables were splitted in 2 levels using the median as the cutoff point. A 2-tailed P value was reported in all instances.

__<a href="#OR">Logistic Regression Analysis.</a>__ Odds ratios (OR) with 95% confidence intervals (CI) and their associated P values were estimated for superficial high-grade vs. deep low-grade tumors, considering inguinal lymph node metastasis, tumor relapse, final nodal status, and cancer-related death.

```{r FiguresGlobal}
library(knitr)
opts_chunk$set(fig.width = 8, fig.height = 6, message = FALSE, warning = FALSE)
```

### <a name="Descriptive">Descriptive Statistics</a>
Here it follows the description of all the variables included in the analyzed dataset.

#### Surgical Procedure for Primary Treatment
```{r Procedure}
Var <- Data$Procedure
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Histologic Subtype
```{r Subtype}
Var <- Data$Subtype
categorical.plot(Var, align = "h", left =8)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Anatomical Location
```{r Anatomical}
Var <- Data$Anatomical
categorical.plot(Var, align = "h", left = 14)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Anatomical Level of Maximum Tumor Invasion
```{r Level}
Var <- Data$Level
categorical.plot(Var, align = "h", left = 9)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Tumor Size, cm
```{r Size}
Var <- Data$Size
numerical.plot(Var, label = "Tumor Size, cm")
descriptive.numerical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Patient's Age, years
```{r Age}
Var <- Data$Age
numerical.plot(Var, label = "Patient's Age, years")
descriptive.numerical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Bilateral Inguinal Lymphadenectomy
```{r Lymphadenectomy}
Var <- Data$Lymphadenectomy
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Inguinal Lymph Node Metastasis
```{r Mets}
Var <- Data$Mets
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Tumor Invasion of Penile Urethra
```{r Urethra}
Var <- Data$Urethra
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Vascular Invasion
```{r Vascular}
Var <- Data$Vascular
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Perineural Invasion
```{r Perineural}
Var <- Data$Perineural
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Pathological T Stage
```{r pT}
Var <- Data$pT
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Tumor Relapse (Local, Regional or Systemic)
```{r Relapse}
Var <- Data$Relapse
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Follow-up Length, months
```{r FollowUp}
Var <- Data$FollowUp
numerical.plot(Var, label = "Follow-Up, Months")
descriptive.numerical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Patient's Outcome
```{r Outcome}
Var <- Data$Outcome
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Clinical N Stage
```{r cN}
Var <- Data$cN
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Final Nodal Status
```{r Final_Nodal}
Var <- Data$Final_Nodal
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

#### Cancer-Related Death
```{r DOD}
Var <- Data$DOD
categorical.plot(Var)
descriptive.categorical(Var)
```

_Number of missing cases: `r sum(is.na(Var))` cases._

***

### <a name="Inferential">Inferential Statistics</a>
```{r Paradoxical_Var}
Var2 <- Data$Paradoxical
```

#### Tumor Type and Surgical Procedure
```{r Paradoxical_Procedure}
Var1 <- Data$Procedure
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Anatomical Location
```{r Paradoxical_Anatomical}
Var1 <- Data$Anatomical
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Anatomical Level
```{r Paradoxical_Level}
Var1 <- Data$Level
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Tumor Size
```{r Paradoxical_Size}
Var1 <- Data$Size
numerical.group.plot(Var1, Var2)
descriptive.numerical.group(Var1, Var2)
```

***

#### Tumor Type and Patient's Age
```{r Paradoxical_Age}
Var1 <- Data$Age
numerical.group.plot(Var1, Var2)
descriptive.numerical.group(Var1, Var2)
```

***

#### Tumor Type and Urethral Invasion
```{r Paradoxical_Urethra}
Var1 <- Data$Urethra
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Vascular Invasion
```{r Paradoxical_Vascular}
Var1 <- Data$Vascular
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Perineural Invasion
```{r Paradoxical_Perineural}
Var1 <- Data$Perineural
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and cN Stage
```{r Paradoxical_cN}
Var1 <- Data$cN
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and pT Stage
```{r Paradoxical_pT}
Var1 <- Data$pT
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Inguinal LN Metastasis
```{r Paradoxical_Mets}
Var1 <- Data$Mets
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Tumor Relapse
```{r Paradoxical_Relapse}
Var1 <- Data$Relapse
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Final Nodal Status
```{r Paradoxical_FN}
Var1 <- Data$Final_Nodal
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

#### Tumor Type and Cancer-Related Death
```{r Paradoxical_DOD}
Var1 <- Data$DOD
categorical.group.plot(Var1, Var2)
descriptive.categorical.group(Var1, Var2)
```

***

### <a name="Survival_FN">Survival Analysis: Final Nodal Status</a>
```{r FN_Survival}
x.lab <- "Follow-Up, Months"
y.lab <- "Survival Function"
# Defining outcome variable
Status <- Data$Final_Nodal
# Creating dicotomic variables from numerical variables for plotting
Size_Median <- factor(ifelse(Data$Size > median(Data$Size, na.rm = TRUE),
  c("Above Median Size"), c("Below Median Size")))
Age_Median <- factor(ifelse(Data$Age > median(Data$Age, na.rm = TRUE),
  c("Above Median Age"), c("Below Median Age")))
# By type of tumor
with(Data, survival.plot(Paradoxical, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Tumor Type"))
# By surgical procedure
with(Data, survival.plot(Procedure, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Surgical Procedure"))
# By anatomical location
with(Data, survival.plot(Anatomical, FollowUp, Status,
  xlab = x.lab, ylab = y.lab, ylim = c(0, 1.09),
  title = "Final Nodal Status by Anatomical Location"))
# By anatomical level
with(Data, survival.plot(Level, FollowUp, Status,
  xlab = x.lab, ylab = y.lab, ylim = c(0, 1.09),
  title = "Final Nodal Status by Anatomical Level"))
# By median tumor size
with(Data, survival.plot(Size_Median, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Median Tumor Size"))
# By median patient's age
with(Data, survival.plot(Age_Median, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Median Patient's Age"))
# By urethral invasion
with(Data, survival.plot(Urethra, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Urethral Invasion"))
# By vascular invasion
with(Data, survival.plot(Vascular, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Vascular Invasion"))
# By perineural invasion
with(Data, survival.plot(Perineural, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Perineural Invasion"))
# By pathological T stage
with(Data, survival.plot(pT, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Pathological T Stage"))
# By clinical N stage
with(Data, survival.plot(cN, FollowUp, Status,
  xlab = x.lab, ylab = y.lab,
  title = "Final Nodal Status by Clinical N Stage"))
```

### <a name="Survival_DOD">Survival Analysis: Cancer-Related Death</a>
```{r DOD_Survival}
# Defining outcome variable
Status <- Data$DOD
# Creating dicotomic variables from numerical variables for plotting
Size_Median <- factor(ifelse(Data$Size > median(Data$Size, na.rm = TRUE),
  c("Above Median Size"), c("Below Median Size")))
Age_Median <- factor(ifelse(Data$Age > median(Data$Age, na.rm = TRUE),
  c("Above Median Age"), c("Below Median Age")))
# By type of tumor
with(Data, survival.plot(Paradoxical, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Tumor Type"))
# By surgical procedure
with(Data, survival.plot(Procedure, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Surgical Procedure"))
# By anatomical location
with(Data, survival.plot(Anatomical, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Anatomical Location"))
# By anatomical level
with(Data, survival.plot(Level, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Anatomical Level"))
# By median tumor size
with(Data, survival.plot(Size_Median, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Median Tumor Size"))
# By median patient's age
with(Data, survival.plot(Age_Median, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Median Patient's Age"))
# By urethral invasion
with(Data, survival.plot(Urethra, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Urethral Invasion"))
# By vascular invasion
with(Data, survival.plot(Vascular, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Vascular Invasion"))
# By perineural invasion
with(Data, survival.plot(Perineural, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Perineural Invasion"))
# By pathological T stage
with(Data, survival.plot(pT, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Pathological T Stage"))
# By clinical N stage
with(Data, survival.plot(cN, FollowUp, Status,
  ylim = c(0.6, 1), position = "bottomright", xlab = x.lab, ylab = y.lab,
  title = "Cancer-Related Death by Clinical N Stage"))
```

***

### <a name="OR">Logistic Regression Analysis</a>
```{r Mets_Model}
# Creating variables for GLM analysis
Paradoxical_Inv <- factor(Data$Paradoxical,
  levels = c("Deep Low-Grade", "Superficial High-Grade"))
Data$Multicompartment <- factor(ifelse(Data$Anatomical == "Glans alone",
  c("No"), c("Yes")))
Data$CC <- factor(ifelse(Data$Level == "Corpus cavernosum",
  c("Yes"), c("No")))
Data$High_pT <- factor(ifelse(Data$pT == "T3",
  c("Yes"), c("No")))
Data$cN_Positive <- factor(ifelse(Data$cN == "cN0",
  c("No"), c("Yes")))
# Creating list of variables and labels
Predictors <- with(Data, list(
  "Superficial high-grade vs. deep low-grade" = Paradoxical_Inv,
  "Primary treatment" = Procedure,
  "Age > median" = Age_Median,
  "Multicompartment tumor" = Multicompartment,
  "Invasion of corpus cavernosum" = CC,
  "Tumor size > median" = Size_Median,
  "Invasion of penile urethra" = Urethra,
  "Vascular invasion" = Vascular,
  "Perineural invasion" = Perineural,
  "High pT (>pT3)" = High_pT,
  "Positive clinical nodes" = cN_Positive
  ))
Varlabels <- names(Predictors)
```

***

#### Inguinal Lymph Node Metastasis
```{r Mets_OR}
Outcome <- Data$Mets
logistic.table(Outcome, Predictors, Varlabels)
```

*** 

#### Tumor Relapse
```{r Relapse_OR}
Outcome <- Data$Relapse
logistic.table(Outcome, Predictors, Varlabels)
```

*** 

#### Final Nodal Status
```{r FN_OR}
Outcome <- Data$Final_Nodal
logistic.table(Outcome, Predictors, Varlabels)
```

*** 

#### Cancer-Related Death
Tumor median size was removed due to the lack of events in either level.
```{r DOD_OR}
Outcome <- Data$DOD
Predictors <- Predictors[-grep("Tumor size > median", Varlabels)]
Varlabels <- names(Predictors)
logistic.table(Outcome, Predictors, Varlabels)
```

*** 
